---
title: "Loop Structures"
output:
  html_notebook: default
  pdf_document: default
---

## For loop

```{r}
#for (<iterator>) {
#  ## do something
#}
```

```{r}
for(i in 1:10) {
  print(i)
}
```
```{r}
x <- c("a", "b", "c", "d")
```

```{r}
for(i in 1:4) { ## or for (i in 1:length(x)) {
  print(x[i])
}
```

```{r}
for(i in seq_along(x)) {
  print(x[i])  
}
```

```{r}
for(letter in x) { 
  print(letter)
}
```

```{r}
for(letter in x) print(letter)
```

```{r}
for(i in 1:4) print(x[i])
```

```{r}
x <- matrix(1:6, 2, 3)

for(i in seq_len(nrow(x))) {
  for (j in seq_len(ncol(x))) {
    print(x[i, j])
  }  
}
x
```

## While loop

```{r}
count <- 0
while(count < 10) {
  print(count)
  count <- count + 1
}
```

```{r}
z <- 5

while(z >= 3 && z <= 10) {
  print(z)
  
  coin <- rbinom(1, 1, .5)
  
  if(coin == 1) {
    z <- z + 1
  } else {
    z <- z - 1
  }
}
```

```{r}
z <- 5
times <- 0

while(z >= 3 && z <= 10) {
  print(z)
  
  coin <- rbinom(1, 1, .5)
  
  ## random walk
  z <- z + if(coin == 1) 1 else -1
  times <- times + 1
}
times
```

## Repeat loop
```{r}
repeat {
  ## do something
  ## at some point
  break
}  
```

## Next
```{r}
for(i in 1:100) {
  if(i <= 20) {
    ## skip the first 20 iterations
    next
  }
  # do something
}
```


## Looping on the command line
- lapply -> Loop over a list and evaluate a function on each element
- sapply -> Same as `lapply` but try to simplify the result
- apply -> Apply a function over the margins of an array
- tapply -> Apply a function over subsets of a vector
- mapply -> Multivariate version of `lapply`

An auxiliary function `split` is also useful, particularly in conjunction with `lapply`

```{r}
x <- list(a = 1:5, b = rnorm(10))
lapply(x, mean)
```

```{r}
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
lapply(x, mean)
```


```{r}
x <- 1:4
lapply(x, runif)
```

```{r}
x <- 1:4
lapply(x, runif, min = 0, max = 10)
```

```{r}
x <- list(a = matrix(1:4, 2, 2), b = matrix(1:6, 3, 2))
x
```

```{r}
lapply(x, function(m) m[,1])
#lapply(x, function(m){ m[,1] }) ## Note the {}}
```


```{r}
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
sapply(x, mean)
```