---
title: "Loop Structures"
output:
  html_notebook: default
  pdf_document: default
---

## For loop

```{r}
#for (<iterator>) {
#  ## do something
#}
```

```{r}
for(i in 1:10) {
  print(i)
}
```
```{r}
x <- c("a", "b", "c", "d")
```

```{r}
for(i in 1:4) { ## or for (i in 1:length(x)) {
  print(x[i])
}
```

```{r}
for(i in seq_along(x)) {
  print(x[i])  
}
```

```{r}
for(letter in x) { 
  print(letter)
}
```

```{r}
for(letter in x) print(letter)
```

```{r}
for(i in 1:4) print(x[i])
```

```{r}
x <- matrix(1:6, 2, 3)

for(i in seq_len(nrow(x))) {
  for (j in seq_len(ncol(x))) {
    print(x[i, j])
  }  
}
x
```

## While loop

```{r}
count <- 0
while(count < 10) {
  print(count)
  count <- count + 1
}
```

```{r}
z <- 5

while(z >= 3 && z <= 10) {
  print(z)
  
  coin <- rbinom(1, 1, .5)
  
  if(coin == 1) {
    z <- z + 1
  } else {
    z <- z - 1
  }
}
```

```{r}
z <- 5
times <- 0

while(z >= 3 && z <= 10) {
  print(z)
  
  coin <- rbinom(1, 1, .5)
  
  ## random walk
  z <- z + if(coin == 1) 1 else -1
  times <- times + 1
}
times
```

## Repeat loop
```{r}
repeat {
  ## do something
  ## at some point
  break
}  
```

## Next
```{r}
for(i in 1:100) {
  if(i <= 20) {
    ## skip the first 20 iterations
    next
  }
  # do something
}
```


## Looping on the command line
- lapply -> Loop over a list and evaluate a function on each element
- sapply -> Same as `lapply` but try to simplify the result
- apply -> Apply a function over the margins of an array
- tapply -> Apply a function over subsets of a vector
- mapply -> Multivariate version of `lapply`

An auxiliary function `split` is also useful, particularly in conjunction with `lapply`

```{r}
x <- list(a = 1:5, b = rnorm(10))
lapply(x, mean)
#'l' in lapply stands for 'list'
```

```{r}
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
lapply(x, mean)
```


```{r}
x <- 1:4
lapply(x, runif)
```

```{r}
x <- 1:4
lapply(x, runif, min = 0, max = 10)
```

```{r}
x <- list(a = matrix(1:4, 2, 2), b = matrix(1:6, 3, 2))
x
```

```{r}
lapply(x, function(m) m[,1])
#lapply(x, function(m){ m[,1] }) ## Note the {}}
```


```{r}
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
sapply(x, mean)
# s for simplify
```


```{r}
x <- matrix(rnorm(200), 20, 10)
apply(x, 2, mean) ## mean of all columns
print("")
apply(x, 1, sum)  ## sum of all rows
```

```{r}
print("")
rowSums(x) == apply(x, 1, sum)
colSums(x) == apply(x, 2, sum)

rowMeans(x) == apply(x, 1, mean)
colMeans(x) == apply(x, 2, mean)
#rowSums, colSums, rowMeans and colMeans are highly optimezed to perform such task.
```

```{r}
x <- matrix(rnorm(200), 20, 10)
apply(x, 1, quantile, probs = c(.25, .75))
```


```{r}
a <- array(rnorm(2 * 2 * 10), c(2, 2, 10))
apply(a, c(1, 2), mean) ## preserve dim(1, 2)

rowMeans(a, dims = 2)
```


```{r}
list(rep(1, 4), rep(2, 3), rep(3, 2), rep(4, 1))

mapply(rep, 1:4, 4:1)
```



```{r}
noise <- function(n, mean, sd) {
  rnorm(n, mean, sd)
}

noise(5, 1, 2)
```

```{r}
noise(1:5, 1:5, 2)
mapply(noise, 1:5, 5:1, 2)
```


```{r}
x <- c(rnorm(10), runif(10), rnorm(10, 1))
f <- gl(3, 10)
f
tapply(x, f, mean)
tapply(x, f, range)
```


```{r}
x <- c(rnorm(10), runif(10), rnorm(10, 1))
f <- gl(3, 10)
split(x, f)
lapply(split(x, f), mean) ## same as tapply(x, f, mean)
```

```{r}
library(datasets)
head(airquality)
```

```{r}
s <- split(airquality, airquality$Month)
lapply(s, function(x){ colMeans(x[, c("Ozone", "Solar.R", "Wind")]) })
```

```{r}
sapply(s, function(x){ colMeans(x[, c("Ozone", "Solar.R", "Wind")]) })
sapply(s, function(x){ colMeans(x[, c("Ozone", "Solar.R", "Wind")], 
                                na.rm = TRUE) })
```

```{r}
x <- rnorm(10)
f1 <- gl(2, 5)
f2 <- gl(5, 2)
f1
f2
interaction(f1, f2)
```

```{r}
str(split(x, list(f1, f2))) ## implicit call of interaction
```

```{r}
str(split(x, list(f1, f2), drop = TRUE)) ## Droping the empty levels
```



## vapply
```{r}
#vapply(data, function, return.type)
#raise a error if sapply returns somethinf different than return.type
```

